## Exploratory Data Analysis

The first formal step that we will be performing is the sample exploration. This crucial initial part of the analysis consists of an examination of differences and relationships between Quality-Control (QC) metrics of the samples from different groups in order to identify poor-quality samples that must be removed before DEA. After that, the sample variables in the metadata also need to be analyzed and filtered based on the percentage of gene expression variance that they explain for each gene. 


## Quality Control Analysis

First we have to explore and compare the the quality-control metrics of the samples in the different groups given by covariates such as age, sex, pregnancy state, group, plate and flowcell. See **Sample Information** section in chapter 04 for a description of these variables.  

<style>
p.comment {
background-color: #E6E6E6;
padding: 10px;
border: 1px solid black;
margin-left: 25px;
border-radius: 5px;
font-family: sans-serif;
}

</style>

<p class="comment">
❓ ***Why is that relevant?*** As you could imagine, technical and methodological differences in all the steps that were carried out during the experimental stages are potential sources of variation in the quality of the samples. Just imagine all that could have been gone wrong or unexpected while experimenting with mice, during the sampling, in the RNA extraction using different batches, when treating samples in different mediums, when preparing libraries in different plates and sequencing in different flowcells. Moreover, the inherent features of the mice from which the samples come from such as age, tissue, sex and pregnancy state could also affect the samples' metrics if, for example, they were separately analyzed and processed. 
</p>


<p class="comment">
❓***But why do we care about mitochodrial and ribosomal counts as QC metrics?*** In the process of mRNA extraction, either by mRNA enrichment (capturing polyadenilated mRNAs) or rRNA-depletion (removing rRNA), we’d expect to have a low number of ribosomal counts, i.e., counts that map to rDNA, and if we don’t, something must have gone wrong with the procedures. In the case of mitochondrial counts something similar occurs: higher mitochondrial rates will be obtained if the cytoplasmic mRNA capture was deficient or if the transcripts were lost by some technical issue, increasing the proportion of mitochondrial mRNAs. As a result, high <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> mitoRate</span> </mark> and <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> rRNA_rate</span> </mark> imply poor quality in the samples.
</p>


<div class="alert alert-info">
  <strong>Note:</strong> the QC metrics were computed with the unprocessed datasets (neither filtered nor normalized) to preserve the original estimates of the samples.
</div>


### Evaluate QC metrics for groups of samples

Fortunately, we can identify to some extent possible factors that could have influenced on the quality of the samples, as well as isolated samples that are problematic. To do that, we will create boxplots that present the distribution of the samples' metrics separating them by sample variables. 

```{r QC_boxplots,  message=FALSE, warning=FALSE}
library("Hmisc")
library("stringr")
library("cowplot")

## Define QC metrics of interest
qc_metrics <- c("mitoRate", "overallMapRate", "totalAssignedGene", "rRNA_rate", "sum", "detected", "ERCCsumLogErr")

## Define sample variables of interest
sample_variables <- c("Group", "Age", "Sex", "Pregnancy", "plate", "flowcell")


## Function to create boxplots of QC metrics for groups of samples

QC_boxplots <- function(qc_metric, sample_var) {
    ## Define sample colors depending on the sample variable
    if (sample_var == "Group") {
        colors <- c("Control" = "brown2", "Experimental" = "deepskyblue3")
    } else if (sample_var == "Age") {
        colors <- c("Adult" = "slateblue3", "Pup" = "yellow3")
    } else if (sample_var == "Sex") {
        colors <- c("F" = "hotpink1", "M" = "dodgerblue")
    } else if (sample_var == "Pregnancy") {
        colors <- c("Yes" = "darkorchid3", "No" = "darkolivegreen4")
    } else if (sample_var == "plate") {
        colors <- c("Plate1" = "darkorange", "Plate2" = "lightskyblue", "Plate3" = "deeppink1")
    } else if (sample_var == "flowcell") {
        colors <- c(
            "HKCG7DSXX" = "chartreuse2", "HKCMHDSXX" = "magenta", "HKCNKDSXX" = "turquoise3",
            "HKCTMDSXX" = "tomato"
        )
    }

    ## Axis labels
    x_label <- capitalize(sample_var)
    y_label <- str_replace_all(qc_metric, c("_" = ""))

    ## x-axis text angle and position
    if (sample_var == "flowcell") {
        x_axis_angle <- 18
        x_axis_hjust <- 0.5
        x_axis_vjust <- 0.7
        x_axis_size <- 4
    } else {
        x_axis_angle <- 0
        x_axis_hjust <- 0.5
        x_axis_vjust <- 0.5
        x_axis_size <- 6
    }

    ## Extract sample data in colData(rse_gene_filt)
    data <- data.frame(colData(rse_gene_filt))

    ## Sample variable separating samples in x-axis and QC metric in y-axis
    ## (Coloring by sample variable)
    plot <- ggplot(data = data, mapping = aes(x = !!rlang::sym(sample_var), y = !!rlang::sym(qc_metric), color = !!rlang::sym(sample_var))) +
        ## Add violin plots
        geom_violin(alpha = 0, size = 0.4, color = "black", width = 0.7) +
        ## Spread dots
        geom_jitter(width = 0.08, alpha = 0.7, size = 1.3) +
        ## Add boxplots
        geom_boxplot(alpha = 0, size = 0.4, width = 0.1, color = "black") +
        ## Set colors
        scale_color_manual(values = colors) +
        ## Define axis labels
        labs(y = y_label, x = x_label) +
        ## Get rid of the background
        theme_bw() +
        ## Hide legend and define plot margins and size of axis title and text
        theme(
            legend.position = "none",
            plot.margin = unit(c(0.5, 0.4, 0.5, 0.4), "cm"),
            axis.title = element_text(size = 7),
            axis.text = element_text(size = x_axis_size),
            axis.text.x = element_text(angle = x_axis_angle, hjust = x_axis_hjust, vjust = x_axis_vjust)
        )


    return(plot)
}



## Plots of all QC metrics for each sample variable
multiple_QC_boxplots <- function(sample_var) {
    i <- 1
    plots <- list()
    for (qc_metric in qc_metrics) {
        ## Call function to create each individual plot
        plots[[i]] <- QC_boxplots(qc_metric, sample_var)
        i <- i + 1
    }
    ## Arrange multiple plots into a grid
    print(plot_grid(plots[[1]], plots[[2]], plots[[3]], plots[[4]], plots[[5]], plots[[6]], plots[[7]], nrow = 2))
}
```


```{r message=FALSE, warning=FALSE}
multiple_QC_boxplots("Age")
```

Initially, when we separate samples by Age, we can appreciate a clear segregation of adult and pup samples in <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> mitoRate</span> </mark>, with higher mitochondrial rates for adult samples and thus, being lower quality samples than the pup ones. We can also see that pup samples have higher <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> totalAssignedGene</span> </mark>, again being higher quality. The samples are very similar in the rest of the QC metrics.
The former differences must be taken into account because they guide further sample separation by Age, which is necessary to avoid dropping most of the adult samples (that are lower quality) in the **QC-based sample filtering** (see below) and to prevent misinterpreting sample variation given by quality and not by mouse age itself. 

```{r message=FALSE, warning=FALSE}
multiple_QC_boxplots("Sex")
```

With Sex, female and male samples have similar QC metrics but there are some female samples that have high <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> mitoRate</span> </mark> and low <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> totalAssignedGene</span> </mark>, but males don´t. That is consistent with the fact that all males are pups and as we previously observed, pup samples present better QC metrics, but since not all pups are males, some female pups also have smaller <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> mitoRate</span> </mark> and greater <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> totalAssignedGene</span> </mark>. So in this case samples' differences are actually dictated by Age and not by Sex.


```{r message=FALSE, warning=FALSE}
multiple_QC_boxplots("Group")
```

Notably, for Group no evident contrasts are seen in the quality of the samples, which means that both control and exposed samples have similar metrics and therefore the differences between them won't be determined by technical factors but effectively by gene expression changes. 

```{r message=FALSE, warning=FALSE}
multiple_QC_boxplots("Pregnancy")
```

Pregnancy variable is interesting because pregnant dams were obviously females and adults, which we already noted are lower quality. Accordingly to that, pregnant samples present overall higher <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> mitoRate</span> </mark> and lower <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> totalAssignedGene</span> </mark>, <mark style= "background-color: #DFF0FE"> <span style="font-family: monospace"> sum</span> </mark>  (library size) and <mark style= "background-color: #DFF0FE"> <span style="font-family: monospace"> detected</span> </mark> (number of expressed genes) than some samples coming from non-pregnant mice. However, these samples have smaller |<mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> ERCCsumLogErr</span> </mark>|, which means that they had smaller differences between expected and observed concentrations of control transcripts. 

Notwithstanding, we must clarify one more time that as in Sex, the trends observed in Pregnancy are all given by Age: it is Age the variable that fully segregates samples in <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> mitoRate</span> </mark> and almost completely in </mark>, lower <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> totalAssignedGene</span> </mark>. Because all pregnant dams were adults, their metrics will be positioned where the adult samples were, but there were also not-pregnant adults that share similar QC values. 


```{r message=FALSE, warning=FALSE}
multiple_QC_boxplots("plate")
```

In plate, no alarming differences are presented but some samples from the 1st plate
have low (though not much lower) <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> overallMapRate</span> </mark>,  <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> totalAssignedGene</span> </mark> and <mark style= "background-color: #DFF0FE"> <span style="font-family: monospace"> sum</span> </mark>. 

```{r message=FALSE, warning=FALSE}
multiple_QC_boxplots("flowcell")
```

For the flowcell, again no worrying distinctions are seen, with the exception of a few individual samples far from the rest. 

In all the previous plots we can appreciate a group of samples placed below in <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> mitoRate</span> </mark> that correspond to pup samples. The relationship is more fuzzy in <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> totalAssignedGene</span> </mark>.


### Examine relationships between QC metrics of samples

Secondly, we want to analyze if there are relationships between QC variables that explain the presence of samples with two or more particular QC metrics. For that, scatterplots are useful because they enable the evaluation of pairs of samples' QC metrics. 

```{r "QC scatterplots", message=FALSE, warning=FALSE}
## Scatterplots for a pair of QC metrics

QC_scatterplots <- function(sample_var, qc_metric1, qc_metric2) {
    ## Define sample colors
    if (sample_var == "Group") {
        colors <- c("Control" = "brown2", "Experimental" = "deepskyblue3")
    } else if (sample_var == "Age") {
        colors <- c("Adult" = "slateblue3", "Pup" = "yellow3")
    } else if (sample_var == "Sex") {
        colors <- c("F" = "hotpink1", "M" = "dodgerblue")
    } else if (sample_var == "Pregnancy") {
        colors <- c("Yes" = "darkorchid3", "No" = "darkolivegreen4")
    } else if (sample_var == "plate") {
        colors <- c("Plate1" = "darkorange", "Plate2" = "lightskyblue", "Plate3" = "deeppink1")
    } else if (sample_var == "flowcell") {
        colors <- c(
            "HKCG7DSXX" = "chartreuse2", "HKCMHDSXX" = "magenta", "HKCNKDSXX" = "turquoise3",
            "HKCTMDSXX" = "tomato"
        )
    }

    data <- colData(rse_gene_filt)

    ## Scatterplots for continuous variable vs continuous variable
    ## First QC metric in x-axis and second QC metric in y-axis
    plot <- ggplot(as.data.frame(data), aes(
        x = !!rlang::sym(qc_metric1),
        y = !!rlang::sym(qc_metric2),
        ## Color samples by a variable
        color = !!rlang::sym(sample_var)
    )) +
        ## Add scatterplot
        geom_point(size = 1) +
        ## Add regression line
        stat_smooth(geom = "line", alpha = 0.4, size = 0.4, span = 0.25, method = lm, color = "orangered3") +
        ## Colors
        scale_color_manual(name = sample_var, values = colors) +
        theme_bw() +
        ## Add Pearson correlation coefficient between the metrics as subtitle
        labs(
            subtitle = paste0("Corr: ", signif(cor(data[, qc_metric1], data[, qc_metric2], method = "pearson"), digits = 3)),
            ## Add axis labels
            y = gsub("_", " ", qc_metric2),
            x = gsub("_", " ", qc_metric1)
        ) +
        ## Plot margins and text size
        theme(
            plot.margin = unit(c(0.1, 1.2, 0.1, 1.2), "cm"),
            axis.title = element_text(size = (7)),
            axis.text = element_text(size = (6)),
            plot.subtitle = element_text(size = 7, color = "gray40"),
            legend.text = element_text(size = 6),
            legend.title = element_text(size = 7)
        )
    return(plot)
}



## QC scatterplots coloring by all sample variables
multiple_QC_scatterplots <- function(qc_metric1, qc_metric2) {
    sample_variables <- c("Age", "Sex", "plate", "Pregnancy", "Group", "flowcell")

    i <- 1
    plots <- list()
    for (sample_var in sample_variables) {
        plots[[i]] <- QC_scatterplots(sample_var, qc_metric1, qc_metric2)
        i <- i + 1
    }
    plot_grid(plots[[1]], plots[[2]], plots[[3]], plots[[4]], plots[[5]], plots[[6]], nrow = 3, rel_widths = c(1, 1))
}
```

```{r message=FALSE, warning=FALSE}
multiple_QC_scatterplots("mitoRate", "rRNA_rate")
```

For <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> mitoRate</span> </mark> vs <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> rRNA_rate</span> </mark> plots there's a negligible positive correlation which was anticipated since no obvious relationship exists between them, though if cytoplasmic mRNAs weren't well captured (lowering the total number of mapping reads) we could expect an increase in both mitochondrial and ribosomal rates. 

Note the complete separation of samples in <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> mitoRate</span> </mark> (but not in <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> rRNA_rate</span> </mark>) by Age, which in turn causes the grouping of male samples (all pups). 


```{r message=FALSE, warning=FALSE}
multiple_QC_scatterplots("mitoRate", "totalAssignedGene")
```

Forcefully, <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> mitoRate</span> </mark>  and <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> totalAssignedGene</span> </mark> have a strong negative correlation, which reveals the existence of very bad quality samples (with small <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> totalAssignedGene</span> </mark> and high <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> mitoRate</span> </mark>).

```{r message=FALSE, warning=FALSE}
multiple_QC_scatterplots("sum", "detected")
```

Noticeably, a positive correlation is present for <mark style= "background-color: #DFF0FE"> <span style="font-family: monospace"> sum</span> </mark> and <mark style= "background-color: #DFF0FE"> <span style="font-family: monospace"> detected</span> </mark> and that is also congruent because a higher number of expressed-genes implies bigger library sizes, but that is not necessarily true in the other way around: just a few highly expressed genes (small number of <mark style= "background-color: #DFF0FE"> <span style="font-family: monospace"> detected</span> </mark> genes) could increase the library size. Interestingly, in this case samples are diagonally separated by age. 


```{r message=FALSE, warning=FALSE}
multiple_QC_scatterplots("sum", "totalAssignedGene")
```

Contrary to expectations, <mark style= "background-color: #DFF0FE"> <span style="font-family: monospace"> sum</span> </mark> is not correlated to <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> totalAssignedGene</span> </mark> and this is unexpected because higher proportions of genes' reads could initially increase library size. However, since <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> totalAssignedGene</span> </mark> is a fraction of the total reads, it could be that a high fraction of reads mapping to genes reflects a small number of total reads and does not necessarily imply more expressed genes (and therefore bigger libraries). In other words, even when <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> totalAssignedGene</span> </mark> equals 1, if the total number of reads is small, the genes won't be widely covered by them and the libraries won't be sizeable. Observe the separarion of samples by Age. 


```{r message=FALSE, warning=FALSE}
multiple_QC_scatterplots("detected", "totalAssignedGene")
```

Lastly, there's a slight positive correlation between <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> totalAssignedGene</span> </mark> and <mark style= "background-color: #DFF0FE"> <span style="font-family: monospace"> detected</span> </mark>, which makes sense because reads assigned to genes rise the number of non-zero expressed genes but as mentioned, those genes are not necessarily highly expressed and thus, library sizes are not perforce so much bigger. Also note that the correlation wouldn't be conserved within pup samples. 


<style>
p.exercise  {
background-color: #E4EDE2;
padding: 10px;
border: 1px solid black;
border-radius: 10px;
font-family: sans-serif;
}

</style>

<p class="exercise">
📑 **Exercise 1**: What's the correlation between <mark style= "background-color: #DFF0FE"> <span style="font-family: monospace"> subsets_Mito_sum</span> </mark> and <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> mitoMapped</span> </mark>? What factor could explain that correlation? Why are <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> mitoMapped</span> </mark> numbers bigger than <mark style= "background-color: #DFF0FE"> <span style="font-family: monospace"> subsets_Mito_sum</span> </mark> values?
</p>

```{r exercise1_EDA, message=FALSE, warning=FALSE,  eval=FALSE, echo=FALSE}
## Solution
multiple_QC_scatterplots("subsets_Mito_sum", "mitoMapped")
## Because in mitoMapped you take reads that mapped to the whole mt chr, in subsets_Mito_sum only reads that were aligned to mt genes. But there's almost a perfect correlation between these two metrics.
```

Remarkably, samples were not differentiated by any other variable (including Group!) but the differences between adult and pup metrics suggest that they must be splitted into two different groups for the following analyses.


### QC sample filtering

After assessing how different or similar are the QC values between samples, we can now proceed to sample filtering based precisely, on these metrics. For that, we will use `isOutlier()` from `r BiocStyle::Biocpkg("scater")` to identify outlier samples only at the lower end or the higher end, depending on the QC metric.

```{r "QC sample filtering", message=FALSE, warning=FALSE}
library("scater")
library("rlang")
library("ggrepel")

## Separate data by Age
rse_gene_pups <- rse_gene_filt[, which(rse_gene_filt$Age == "Pup")]
rse_gene_adults <- rse_gene_filt[, which(rse_gene_filt$Age == "Adult")]


## Find outlier samples based on their QC metrics (samples that are 3 median-absolute-deviations away from the median)


## Filter all samples together

## Drop samples with lower library sizes (sum), detected number of genes and totalAssignedGene
outliers_library_size <- isOutlier(rse_gene_filt$sum, nmads = 3, type = "lower")
outliers_detected_num <- isOutlier(rse_gene_filt$detected, nmads = 3, type = "lower")
outliers_totalAssignedGene <- isOutlier(rse_gene_filt$totalAssignedGene, nmads = 3, type = "lower")
## Drop samples with higher mitoRates and rRNA rates
outliers_mito <- isOutlier(rse_gene_filt$mitoRate, nmads = 3, type = "higher")
outliers_rRNArate <- isOutlier(rse_gene_filt$rRNA_rate, nmads = 3, type = "higher")

## Keep not outlier samples
not_outliers <- which(!(outliers_library_size | outliers_detected_num | outliers_totalAssignedGene | outliers_mito | outliers_rRNArate))
rse_gene_qc <- rse_gene_filt[, not_outliers]

## Number of samples retained
dim(rse_gene_qc)[2]

## Add new variables to rse_gene_filt with info of samples retained/dropped
rse_gene_filt$Retention_after_QC_filtering <- as.vector(sapply(rse_gene_filt$SAMPLE_ID, function(x) {
    if (x %in% rse_gene_qc$SAMPLE_ID) {
        "Retained"
    } else {
        "Dropped"
    }
}))



## Filter adult samples

outliers_library_size <- isOutlier(rse_gene_adults$sum, nmads = 3, type = "lower")
outliers_detected_num <- isOutlier(rse_gene_adults$detected, nmads = 3, type = "lower")
outliers_totalAssignedGene <- isOutlier(rse_gene_adults$totalAssignedGene, nmads = 3, type = "lower")
outliers_mito <- isOutlier(rse_gene_adults$mitoRate, nmads = 3, type = "higher")
outliers_rRNArate <- isOutlier(rse_gene_adults$rRNA_rate, nmads = 3, type = "higher")

not_outliers <- which(!(outliers_library_size | outliers_detected_num | outliers_totalAssignedGene | outliers_mito | outliers_rRNArate))
rse_gene_adults_qc <- rse_gene_adults[, not_outliers]

## Number of samples retained
dim(rse_gene_adults_qc)[2]

rse_gene_adults$Retention_after_QC_filtering <- as.vector(sapply(rse_gene_adults$SAMPLE_ID, function(x) {
    if (x %in% rse_gene_adults_qc$SAMPLE_ID) {
        "Retained"
    } else {
        "Dropped"
    }
}))



## Filter pup samples

outliers_library_size <- isOutlier(rse_gene_pups$sum, nmads = 3, type = "lower")
outliers_detected_num <- isOutlier(rse_gene_pups$detected, nmads = 3, type = "lower")
outliers_totalAssignedGene <- isOutlier(rse_gene_pups$totalAssignedGene, nmads = 3, type = "lower")
outliers_mito <- isOutlier(rse_gene_pups$mitoRate, nmads = 3, type = "higher")
outliers_rRNArate <- isOutlier(rse_gene_pups$rRNA_rate, nmads = 3, type = "higher")

not_outliers <- which(!(outliers_library_size | outliers_detected_num | outliers_totalAssignedGene | outliers_mito | outliers_rRNArate))
rse_gene_pups_qc <- rse_gene_pups[, not_outliers]

## Number of samples retained
dim(rse_gene_pups_qc)[2]

rse_gene_pups$Retention_after_QC_filtering <- as.vector(sapply(rse_gene_pups$SAMPLE_ID, function(x) {
    if (x %in% rse_gene_pups_qc$SAMPLE_ID) {
        "Retained"
    } else {
        "Dropped"
    }
}))
```

We already filtered outlier samples ... but what have we removed?
It is always important to trace the QC metrics of the filtered samples to verify that they really are poor quality. We don't want to get rid of useful samples! So let's go back to the QC boxplots but now color samples according to whether they passed or not the filtering step and also distinguishing samples' groups by shape. 

```{r message=FALSE, warning=FALSE}
## Boxplots of QC metrics after sample filtering

## Boxplots
boxplots_after_QC_filtering <- function(rse_gene, qc_metric, sample_var) {
    ## Color samples
    colors <- c("Retained" = "deepskyblue", "Dropped" = "brown2")

    ## Sample shape by sample variables
    if (sample_var == "Group") {
        shapes <- c("Control" = 0, "Experimental" = 15)
    } else if (sample_var == "Age") {
        shapes <- c("Adult" = 16, "Pup" = 1)
    } else if (sample_var == "Sex") {
        shapes <- c("F" = 11, "M" = 19)
    } else if (sample_var == "Pregnancy") {
        shapes <- c("Yes" = 10, "No" = 1)
    } else if (sample_var == "plate") {
        shapes <- c("Plate1" = 12, "Plate2" = 5, "Plate3" = 4)
    } else if (sample_var == "flowcell") {
        shapes <- c(
            "HKCG7DSXX" = 3, "HKCMHDSXX" = 8, "HKCNKDSXX" = 14,
            "HKCTMDSXX" = 17
        )
    }


    y_label <- str_replace_all(qc_metric, c("_" = " "))

    data <- data.frame(colData(rse_gene))

    ## Median of the QC var values
    median <- median(eval(parse_expr(paste("rse_gene$", qc_metric, sep = ""))))
    ## Median-absolute-deviation of the QC var values
    mad <- mad(eval(parse_expr(paste("rse_gene$", qc_metric, sep = ""))))

    plot <- ggplot(data = data, mapping = aes(
        x = "", y = !!rlang::sym(qc_metric),
        color = !!rlang::sym("Retention_after_QC_filtering")
    )) +
        geom_jitter(alpha = 1, size = 2, aes(shape = eval(parse_expr((sample_var))))) +
        geom_boxplot(alpha = 0, size = 0.15, color = "black") +
        scale_color_manual(values = colors) +
        scale_shape_manual(values = shapes) +
        labs(x = "", y = y_label, color = "Retention after QC filtering", shape = sample_var) +
        theme_classic() +
        ## Median line
        geom_hline(yintercept = median, size = 0.5) +
        ## Line of median + 3 MADs
        geom_hline(yintercept = median + (3 * mad), size = 0.5, linetype = 2) +
        ## Line of median - 3 MADs
        geom_hline(yintercept = median - (3 * mad), size = 0.5, linetype = 2) +
        theme(
            axis.title = element_text(size = (9)),
            axis.text = element_text(size = (8)),
            legend.position = "right",
            legend.text = element_text(size = 8),
            legend.title = element_text(size = 9)
        )

    return(plot)
}
```

In the following plot we can confirm that taking all samples together (both adults and pups) all adult samples are dropped by their high <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> mitoRate</span> </mark>, but that is not desirable because we want to analyze these samples too, so we need to analyze pup and adult samples separately. 

```{r message=FALSE, warning=FALSE}
## Plots

## All samples together
p <- boxplots_after_QC_filtering(rse_gene_filt, "mitoRate", "Age")
p + theme(plot.margin = unit(c(2, 4, 2, 4), "cm"))
```

For adults, only 3 controls are dropped, 3 of them by their <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> mitoRate</span> </mark>. 

```{r message=FALSE, warning=FALSE}
## Adult samples
p <- boxplots_after_QC_filtering(rse_gene_adults, "mitoRate", "Group")
p + theme(plot.margin = unit(c(2, 4, 2, 4), "cm"))
```

For pups, 1 experimental sample was dropped by its <mark style= "background-color: #FCF3CF"> <span style="font-family: monospace"> rRNA_rate</span> </mark>.

```{r message=FALSE, warning=FALSE}
## Pup samples
p <- boxplots_after_QC_filtering(rse_gene_pups, "rRNA_rate", "Group")
p + theme(plot.margin = unit(c(2, 4, 2, 4), "cm"))
```

<p class="exercise">
📑 **Exercise 2**: Were the filtered adult samples from pregnant or not pregnant mice? From which plates and flowcells? What were their library sizes?
</p>

```{r exercise2_EDA, message=FALSE, warning=FALSE, eval=FALSE, echo=FALSE}
## Solution
p <- boxplots_after_QC_filtering(rse_gene_adults, "mitoRate", "Pregnancy")
p + theme(plot.margin = unit(c(2, 4, 2, 4), "cm"))

p <- boxplots_after_QC_filtering(rse_gene_adults, "mitoRate", "plate")
p + theme(plot.margin = unit(c(2, 4, 2, 4), "cm"))

p <- boxplots_after_QC_filtering(rse_gene_adults, "mitoRate", "flowcell")
p + theme(plot.margin = unit(c(2, 4, 2, 4), "cm"))

p <- boxplots_after_QC_filtering(rse_gene_adults, "sum", "Group")
p + theme(plot.margin = unit(c(2, 4, 2, 4), "cm"))
```

<div class="alert alert-info">
   Try the function yourself with different QC metrics and sample variables! 
</div>



## References {-}

1. Evans, C., Hardin, J., & Stoebel, D. M. (2018). Selecting between-sample RNA-Seq normalization methods from the perspective of their assumptions. *Briefings in bioinformatics*, 19(5), 776-792.

2. Bedre, R. (2023). Gene expression units explained: RPM, RPKM, FPKM, TPM, *DESeq*, TMM, SCnorm, GeTMM, and ComBat-Seq. Web site: https://www.reneshbedre.com/blog/expression_units.html
